Paper 1: Triangulating trends in inequality around the world from secondary databases

### 2023-05-02 PhD_pages/Paper_1/paper1.qmd
Paper 1: Drafting

@@

Economic inequalities around the world are vast. The world's richest 1% enjoy an average income that is around 150 times that of the poorer half of the world's population [@chancel_world_2021]. While the richest may afford private flights to space, roughly three billion people are unable to afford even just the cost of a healthy diet [@herforth_cost_2020].

@@

Their use and impact, however, has taken place within largely separate arenas. Many methodological differences separate the two datasets, including the  underlying data sources they rely on, the methods and assumptions applied to those sources, the concept of welfare measured, the summary indices of inequality typically favoured, and coverage they provide. The two datasets represent distinct and, to a large extent, incommensurable 'worlds' of inequality measurement. This lack of alignment between the datasets limits our ability to combine the insights provided by each into a coherent empirical view informed by both.

@@

The two datasets have also yielded starkly contrasting accounts of how big a contribution within-country inequalities make to overall global inequality. Most analyses of global income inequality to date have relied heavily on the *PIP* data [@kanbur_global_2022; @lakner_global_2016; @anand_global_2015; @milanovic_global_2015; @Nino-Zarazua2017-tv]. A consistent result from these studies is that, despite some recent convergence in average incomes between countries [@patel_new_2021], such differences  continue to make up the lion's share of global inequality. On such accounts, a person's income is largely determined purely by the country in which they happen to have been born [@milanovic_global_2015]. More recent analysis of the *WID* data, however, flips this established result on its head. @chancel_global_2021 find within-country inequalities instead 'dominate' inequality between countries, contributing 78% in a decompositon of the global Theil index.

The two worlds of *measurement* thus depict two very different empirical *realities*. Knowing which of these worlds we live in matters not just for our understanding of inequality today, but also our understanding of how we should seek to combat inequality in the future. A near ubiquitous rise in inequality apparent in the *WID* data points to fundamental 'forces for divergence' within capitalism [@piketty_capital_2014] and the need for global coordination, particularly on the taxation of capital, to counter them [@saez_triumph_2019]. If inequalities within countries now predominate within the global distribution of income, such avenues will also be central to tackling global inequality too [@piketty_brief_2022]. Conversely, the much more mixed picture seen in the *PIP* data highlights more prominently the past role and future scope of national policies to shape the extent of inequality within countries [@ravallion_inequality_2018]. From this viewpoint, such inequalities in any case remain a small share of overall global inequality such that the global distribution will continue to depend heavily on future aggregate growth rates [@kanbur_global_2022].

More fundamentally, the two datasets support differing narratives concerning globalization. [TODO: elabourate].

Despite laudable efforts to make transparent, in practice a black box. Treatmetn of the underlying data and assumptions. Whilst strengths and weaknesses studied for particular countries with good data, when applied globally we really don't know. Despite the great data, fundamental ambiguity surrounding the most basic aspects.

This is the first of a pair of papers that try to address this problem by bridging between the *PIP* and *WID* datasets, and the two 'worlds' of measurement they manifest. The novelty of the papers lies preciesly in considering both datasets in conjunction, in contrast to the many studies that consider only one in isolation. In taking such a combined view, we are able to 'triangulate' a more robust picture of inequality around the world by highlighting where the datasets agree and reinforce one another, and pinpointing more precisely where and why they do not.

The present paper the focuses on trends in within-country inequality. A companion paper considers overall global inequality, factoring in differences in average incomes between countries alongside the within-country inequalities analysed here.

The rest of the paper is organized as follows. Section 2 provides a more detailed description of the *PIP* and *WID* datasets and their main differences. Section 3 compares within-country inequality trends in the two datasets, starting with the metrics where they fall furthest apart. For this we use the Gini of disposabale income (or consumption) and the top 1% share of pretax income as metrics emblematic of the *PIP* and *WID* datasets respectively.
Section 4 takes a first simple step to bridge betweent he datasets, by aligning the coverage. Section 5 considers a range of inequality indices, comparing the datasets on a like-for-like basis.
Section 5 zooms in subset of countries where there is most disagreement. the role of welfare metrics and specific data quality issues for 
Section 6 concludes, drawing out the main overlaps and disagreements between the two datasets, accounting for the steps to

>>>>

# On such accounts, a person's income is determined largely simply by the country in which they happen to have been born [@milanovic_global_2015].

# A near ubiquitous rise in inequality apparent in the *WID* data points to fundamental 'forces for divergence' within capitalism [@piketty_capital_2014] and the need for global coordination, particularly on the taxation of capital, to counter them [@saez_triumph_2019]. If inequalities within countries now predominate within the global distribution of income, such avenues will also be central to tackling global inequality too [@piketty_brief_2022]. Conversely, the much more mixed picture seen in the *PIP* data highlights more prominently the past role and future scope of national policies to shape the extent of inequality within countries [@ravallion_inequality_2018]. From this viewpoint, such inequalities in any case remain a small share of overall global inequality such that the global distribution will continue to depend heavily on future aggregate growth rates [@kanbur_global_2022].

@@

Economic inequalities around the world are vast. While the richest can afford private flights to space, roughly three billion people are unable to afford even just the cost of a healthy diet [@herforth_cost_2020]. The world's richest 1% enjoy an average income that is around 150 times that of those in the poorer half of the world's population [@chancel_world_2021].

@@

Their use and impact, however, has taken place within largely separate arenas. Many methodological differences distinguish the *PIP* and *WID* datasets. This includes the underlying data sources they rely on; the methods and assumptions applied to those sources; the concept of welfare measured; the summary indices of inequality typically favoured; and coverage they provide. This lack of alignment makes it difficult to combine the insights each provide into a coherent empirical view informed by both. The two datasets represent distinct and, to a large extent, incommensurable 'worlds' of inequality measurement – used in isolation rather than in conjunction.

@@

The two datasets have also yielded opposing accounts of the contribution made by within-country inequalities to overall global inequality. Most analyses of global income inequality to date have relied heavily on the *PIP* data [@kanbur_global_2022; @lakner_global_2016; @anand_global_2015; @milanovic_global_2015; @Nino-Zarazua2017-tv]. A consistent result from these studies is that differences between countries' average incomes continue to make up the lion's share of global inequality, despite some recent convergence [@patel_new_2021]. More recent analysis of the *WID* data, however, flips this established result on its head. @chancel_global_2021 find within-country inequalities instead now 'dominate' inequality between countries, contributing 78% in a decompositon of the global Theil index.

Alongside these divergent assesments of *how* inequality has changed around the world, the two approaches to measurement are frequently used to evidence contrasting accounts of *why* it has changed in this way and hence where to target efforts to combat inequality in the future. [TODO: Expand thought: i.e. marginal labour productivies/wage distribution vs capital incomes/fundamental force for divergence/top marginal taxes/top wages bargaining.].

The two 'worlds' of inequality measurement adopted by the *PIP* and *WID* datasets have yielded two very different views of inequality around the world and consequently two very different sets of recommendations as to how to address it.

Which of these realities should we give most credence to? A simplistic reading of the top incomes literature would see the *WID* data as unambiguously providing greater reliability, thanks to the incorporation of tax data and other sources in order to correct for the well-known issue of the 'missing rich' within survey data [@lustig_missing_2019]. The need for such a correction is amply demonstrated in many cases. @korinek_survey_2006, for example, find richer households in the United States to be less likely to participate in surveys and estimate this to bias the Gini coefficient downwards by around five percentage points. [TODO: More examples, including underreporting]. The *PIP* data, based purely on survey data, makes no such correction and is thus likely to understimate the extent of inequality in many cases.

A more agnostic assessment of the two datasets is, however, warrented on a number of grounds. Firstly, the extent of the upward revision to inequality estimates resulting from top income adjustments varies from country to country. In some cases it may be quite modest, as demonstrated by @carranza_top_2021. They apply *WID*-style top income adjustments to survey data for 26 European countries and find an average increase in the Gini of around 2.4 points, with notable differences across countries and limited impact on trends. Secondly, for many countries/periods included in the *WID* estimates, the richer range of data sources that the dataset's methodology seeks to incorporate do not exist. In the place of data, various imputation and interpolation strategies must be employed.^[For example, see the methods outlined in @chancel_towards_2019 in the context of Africa, for which data quality and coverage is especially low] The terms of this tradeoff are not clear. While the incorporation of tax data is a proven asset in many individual cases, the value of top income adjustment methodologies when applied *globally* is harder to demonstrate. Lastly, deviations between survey and tax data appear to be restricted to the *very* top of the distribution. After aligning the income concept across survey and tax data for United States, Germany and France, @yonzan_drawing_2022 find that the sources correspond extremely well up until the top 1% of the distribution. The apparent 'accuracy' of survey data is thus likely to depend on the inequality index chosen, and in particular to its sensitity to changes at the top of the distribution. It is thus a question whose answer will depend also on normative judgements concerning the nature of inequality [@atkinson_measurement_1970].

Overall, while for particular questions and cases, much can be said about the relative strengths and weaknesses of the two approaches to measurement employed in the *PIP* and *WID* datasets, there is no straightforward answer to the question of which offers us the most reliable or relevant global view on inequality.

A different route, however, may be taken to try to reconcile the divergent views of inequality drawn from the two datasets. A limitation of the existing .tend to be made insolation. This means areas of agreement are overlooked and the reasons for disagreement are obfuscated. By using the two datasets in conjunction, rather than in isolation, we can 'triangulate' a more robust and coherent empirical view informed by both. [TODO: Add footnote citing Tony Atkinson on the idea of triangulation]. This is the first of a pair of papers that seeks to do this. The present paper focuses on trends in within-country inequality – a companion paper takes a similar approach but considers overall global inequality. It triangulates between the *PIP* and *WID* data by separately aligning coverage, the inequality index, and where possible the welfare concept.

The result of this is X, Y, Z.

The rest of the paper is organized as follows. Section 2 provides a more detailed description of the *PIP* and *WID* datasets and their main differences. Section 3 compares within-country inequality trends in the two datasets, starting with the metrics they are most associated with and where they fall furthest apart: for *PIP* the Gini of disposabale income (or consumption); for *WID* the top 1% share of pretax national income. Section 4 takes a first simple step to bridge between the datasets, by restricting the analysis to only those countries available in both datasets. Section 5 considers a range of inequality indices, comparing the datasets on a like-for-like basis. Section 5 zooms in subset of countries where there is most disagreement and using Luxembourg Income Study considers the contribution of the different taget welfare concepts employed within the two datasets. Section 6 concludes.
